---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FIA stem map

<!-- badges: start -->
<!-- badges: end -->

The Forest Inventory and Analysis (FIA) Program of USDA Forest Service provide tree-level measurements from a systematic grid of field plots across all forest land uses and ownerships in the US. **FIAstemmap** provides functions for mapping tree stem locations on FIA plots, modeling individual crown dimensions, and generating plot-level estimates of fractional tree canopy cover. Several stand height metrics can be calculated. Spatial analysis of tree point pattern is facilitated for the standard FIA four-point cluster plot design. Efficient data processing supports national applications. This R package provides an updated implementation of the software originally described by Toney et al. 2009 [[1]](#References). Since that original development, the stem-map model for estimating plot canopy cover from tree measurements has supported several applications of FIA data, including:

* reference data for classifying vegetation types and mapping tree canopy cover in the LANDFIRE Program [[2, 3]](#References)
* National Land Cover Database (NLCD) Tree Canopy Cover (TCC) science and development [[4]](#References)
* wildlife habitat analysis [[5]](#References)
* comparative assessments of tree canopy cover estimation methods [[6, 7]](#References)

## Installation

You can install the development version of **FIAstemmap** with:

``` r
# install.packages("pak")
pak::pak("ctoney/FIAstemmap")
```

## Example

TODO...

```{r example}
library(FIAstemmap)
## basic example code
```

## References

[1] Toney, Chris; Shaw, John D.; Nelson, Mark D. 2009. A stem-map model for predicting tree canopy cover of Forest Inventory and Analysis (FIA) plots. In: McWilliams, Will; Moisen, Gretchen; Czaplewski, Ray, comps. _Forest Inventory and Analysis (FIA) Symposium 2008_; October 21-23, 2008; Park City, UT. Proc. RMRS-P-56CD. Fort Collins, CO: U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station. 19 p. Available: https://research.fs.usda.gov/treesearch/33381.

[2] LANDFIRE: LANDFIRE Existing Vegetation Cover layer. (LF2024 version releasd 2025 - last update). U.S. Department of Interior, Geological Survey, and U.S. Department of Agriculture. [Online]. Available: https://landfire.gov/vegetation/evc [accessed 2026, Feb 24].

[3] Moore, Annabelle; La Puma, Inga; Dillon, Greg; Smail, Tobin; Schleeweis, Karen; Toney, Chris; Menakis, Jim; Bastian, Henry; Picotte, Josh; Dockter, Daryn; Tolk, Brian. 2024. Twenty years of science and management with LANDFIRE. Connected Science, October 2024. Fort Collins, CO: U.S. Department of Agriculture, Forest Service, Rocky Mountain Research Station. 2 p. Available: https://research.fs.usda.gov/treesearch/68397.

[4] Toney, Chris; Liknes, Greg; Lister, Andy; Meneguzzo, Dacia. 2012. Assessing alternative measures of tree canopy cover: Photo-interpreted NAIP and ground-based estimates. In: McWilliams, Will; Roesch, Francis A. eds. 2012. _Monitoring Across Borders: 2010 Joint Meeting of the Forest Inventory and Analysis (FIA) Symposium and the Southern Mensurationists_. e-Gen. Tech. Rep. SRS-157. Asheville, NC: U.S. Department of Agriculture, Forest Service, Southern Research Station. 209-215. Available: https://research.fs.usda.gov/treesearch/41009.

[5] Tavernia, B., Nelson, M., Goerndt, M., Walters, B., & Toney, C. (2013). Changes in forest habitat classes under alternative climate and land-use change scenarios in the northeast and midwest, USA. _Mathematical and Computational Forestry & Natural-Resource Sciences_ (MCFNS), **5**:2, 135-150(15). Retrieved from https://www.mcfns.com/index.php/Journal/article/view/MCFNS_165.

[6] Riemann, R., Liknes, G., O’Neil-Dunne, J., Toney, C., Lister, T. (2016). Comparative assessment of methods for estimating tree canopy cover across a rural-to-urban gradient in the mid-Atlantic region of the USA. _Environmental Monitoring and Assessment_ **188**:297. https://doi.org/10.1007/s10661-016-5281-8.

[7] Andrew N. Gray, Anne C.S. McIntosh, Steven L. Garman, Michael A. Shettles. 2021. Predicting canopy cover of diverse forest types from individual tree measurements. _Forest Ecology and Management_, Volume 501, 119682, ISSN 0378-1127, https://doi.org/10.1016/j.foreco.2021.119682.
